material and methods

### 2018-01-03 man/ms_1_material_methods.Rmd
update

### Tree sampling 
Samplig was carried during autumn of 2016. Trees were sampled at two locations in contrasting slopes of Sierra Nevada: San Juan (SJ; northern site) and Cáñar (CA; southern site) (Figure 1; Table 1). Two elevations were sampled for the southern site (CA-Low and CA-High). All the sites were oak monospecific and representatives of two of the three the population's cluster identified for the specie in this mountain range (:red_circle: mejorar; citar Pérz-Luque et al..). In each site between 15 and 20 dominant trees were randomly selected. Two cores of 5 mm of diameter were taken from each at breast heith (1.3 m) using an increment borer. For each tree, diameter at breat height (DBH) and total height were recorded.

Incluir datos de muestreo de competencia

@@

The growth trends were analyzed at different time scales. The long-term growth response was analyzed using basal area increment (hereafter BAI, $\mathrm{cm^2 \cdot year^{-1}}$). BAI represents a more accurate indicator of growth than ring-width, since it removes variation in growth attributatle to increasing stem circumference [@Biondi2008]. We used measured dbh and raw nondetrened ring-widths to compute BAI by subtracting twice the annual ring width from the annual diameter, starting from the measured diameter outside the bark [@Piovesan2008]. We calculated a mean BAI series for each individual tree. 
To study the response of growth to the inter-annual variability of climate (short-term response) we used pre-withened residual chronologies (growth index, GIres) calculated from biweight means of ratios between raw growth measurements and individual cubic splines with a 50 % frecuency cutoff at 30 years [@Fritts1976].

@@

Site chronologies were built by averaging all tree BAI measurement of the same site. To explore similarity within locality, each site chronology was smoothed using centred moving averages with different window sizes, and then Pearson's correlation coefficient between the two chronologies of the same locality (higher and lower elevation) were calculated. Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` [@Canty2016]

>>>>

### Tree sampling 
Samplig was carried during autumn of 2016. Trees were sampled at two locations in contrasting slopes of Sierra Nevada: San Juan (SJ; northern site) and Cáñar (CA; southern site) (Figure 1; Table 1). Two elevations were sampled for the southern site (CA-Low and CA-High). All the sites were oak monospecific and representatives of two of the three the population's cluster identified for the specie in this mountain range (:red_circle: mejorar; citar Pérz-Luque et al..). In each site between 15 and 20 dominant trees were randomly selected. Two cores of 5 mm of diameter were taken from each at breast heith (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured using a girth tape and a Vertex IV ultrasonic hypsometer (Haglöf, Sweden) respectively. We assessed stand competition of the target tree by recording all neighboring living trees with dbh > 7.5 cm in a circle of 10 m radi. The species, dbh and total height were annotated. We also measured distance and azimuth to the target tree. (:red_circle: indices competencia??)

@@

The growth trends were analyzed at different time scales. The long-term growth response was analyzed using basal area increment (hereafter BAI, $\mathrm{cm^2 \cdot year^{-1}}$). BAI represents a more accurate indicator of growth than ring-width, since it removes variation in growth attributatle to increasing stem circumference [@Biondi2008]. We used measured dbh and raw nondetrened ring-widths to compute BAI by subtracting twice the annual ring width from the annual diameter, starting from the measured diameter outside the bark [@Piovesan2008]. We calculated a mean BAI serie for each individual tree. (:red_circle: equation BAI). ... Site chronologies were built by averaging all tree BAI measurement of the same site.

To study the response of growth to the inter-annual variability of climate (short-term response) we used pre-withened residual chronologies calculated from ratios between raw growth measurements and individual cubic splines with a 50 % frecuency cutoff at 30 years [@Fritts1976](:red:circle: (Cook and Kairiukstis 1990).. We standardized and detrended the tree-ring width series using the `dplR` [@Bunn2008, @Bunn2010] package. Residual site chronologies were obtained computing the bi-weight robust mean of all prewhitened growth indices for the trees of the same site.

Statistical descriptive parameters (Fritts, 2001), including the mean, standard deviation, first-order autocorrelation of raw series, the mean sensitivity (a measure of the year-to-year variability) and the mean correlation between individual series of residual ring-width indices, were also calculated for each site chronology considering the common 1950–2013 period.

* incluir tree clim [@Zang2015],  EPS

... The diameters at the breast height (DBH) measured in the field were used to convert each individual ring-width series to basal area increment (BAI) using dplR (Bunn 2008).

The chronologies were computed as bi-weight robust mean to reduce the bias caused by extreme values and an autoregressive model was applied to remove the autocorrelation of the series (Cook and Kairiukstis 1990). The statistical quality of each chronology was checked via expressed population signal (EPS; Wigley et al. 1984). A threshold value of EPS >0.85 was used to determine the cutoff year of the time span that could be considered reliable. The common period 1950–2012 was used for further analyses in order to avoid the period of intensive logging in TJN (see also Fig. 2)....

### Climate and growth

Climate data were obtained from the European Daily High-Resolution Observational Gridded Dataset (E-OBS v16) [@Haylock2008]. Monthly precipitation and mean, minimum and maximum temperatures were obtained at a 0.25 x 0.25 º resolution for the 1950-2016 period. We selected grid cells covering each sampled sites.

treeclim aqui

The SPEI drought index was calculated by applying the R Package “SPEI” [Vicente-Serrano et al., 2010]. The index was calculated for different seasonal windows from 1 to 12 months and at different starting months (from January to December) for the period 1950–2013.

@@

 To explore similarity within locality, each site chronology was smoothed using centred moving averages with different window sizes, and then Pearson's correlation coefficient between the two chronologies of the same locality (higher and lower elevation) were calculated. Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` [@Canty2016]
